---
title: "TO 414 Group Project 1"
author: "Shibi Sanjeev, Bryan Yu, Anthony Zhao, Ben Smith, Jackson Rocke"
date: "2/10/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Group Project 1: Due on 2/21
emails = 
bssmit
tzhx
rockej
bryanyu

```{r}
library(tidyverse)
hotel_data = read.csv("hotel_bookings.csv")

hotel_data$arrival_date_year = as.factor(hotel_data$arrival_date_year)
hotel_data$arrival_date_month = as.factor(hotel_data$arrival_date_month)
hotel_data$arrival_date_week_number = as.factor(hotel_data$arrival_date_week_number)
hotel_data$arrival_date_day_of_month = as.factor(hotel_data$arrival_date_day_of_month)
```

**Basic Data Structure**
```{r}
#str(hotel_data)
```

## **Cancellation Analysis**

An important initial note. We don't have the data for every month in every year. In 2016, we have data for all three months. In 2017, we have data from January until August. In 2015, we only have data from July to December. The only months with three years of data are July and August. All other months only have two years worth of data from either 2015 and 2016 or 2016 and 2017.

**Number of cancellations broken down by year and month**
```{r}
#This new column month_num has the month of the arrival at the hotel, similar to the arrival_date_month column. However, this one is in numerical form. This helps us with graphing. 
hotel_data$month_num = match(hotel_data$arrival_date_month, month.name)

#Number of cancellations broken down by year then month
num_cancellations_by_year_month = group_by(hotel_data, arrival_date_year, month_num, is_canceled) %>%
  summarize(num_cancellations = n())

ggplot(num_cancellations_by_year_month, aes(x = month_num, y = num_cancellations)) +
  geom_bar(aes(fill = as.factor(is_canceled)),stat = "identity") +
  facet_wrap(~arrival_date_year) +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1)) +
  scale_fill_discrete(name = "", labels = c("Not Canceled", "Canceled")) +
  ggtitle("Number of Cancellations and Reservations in Each Year For Each Month") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Number of reservations broken down by year and month but trend lines**
```{r}
#Number of reservations in each month, broken down by year
num_res_year_month = hotel_data %>%
  group_by(arrival_date_year, month_num) %>%
  summarize(num_reservations = n())

ggplot(num_res_year_month, aes(x = month_num, y = num_reservations)) +
  geom_smooth(aes(color = arrival_date_year)) +
  geom_point(aes(color = arrival_date_year)) +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1))
```

**Percentage of reservations canceled broken down by year and month**
```{r}
#filtering data to show how many cancellations in each year and month
test = num_cancellations_by_year_month %>%
  filter(is_canceled == 1) #selecting all the 

percent_reservations_canceled = merge(num_res_year_month, test, by = c("arrival_date_year", "month_num")) %>%
  mutate(percent_canceled = (num_cancellations/num_reservations)*100) %>%
  select(arrival_date_year, month_num, percent_canceled)

ggplot(percent_reservations_canceled, aes(x = month_num, y = percent_canceled)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1)) +
  facet_wrap(~arrival_date_year) +
  ggtitle("Percentage of Cancellations in Each Year For Each Month") +
  theme(plot.title = element_text(hjust = 0.5))
  
```


**Length of Stay**

After calculating the total length of stay, we show the total length of stay with basic descriptive statistics and a histogram. We can see that the mean and median are 3.428 and 3 days respectively.
```{r}
#Total length of stay
hotel_data$total_stay = hotel_data$stays_in_week_nights + hotel_data$stays_in_weekend_nights

summary(hotel_data$total_stay)

filter(hotel_data, total_stay < 20) %>%
ggplot(aes(x = total_stay)) +
  geom_histogram(binwidth = 1) +
  geom_vline(aes(xintercept = mean(total_stay)), color = "lightblue") +
  geom_vline(aes(xintercept = median(total_stay)), color = "navy")
```

**Lead Time and Cancellations**
```{r}
#How many cancelations per days of lead time
group_by(hotel_data, is_canceled, lead_time) %>%
  summarize(count = n())

#Histograms of lead time by whether or not there were cancellations.
filter(hotel_data, lead_time < 600) %>%
ggplot(aes(x = lead_time)) +
  geom_histogram() +
  facet_wrap(~is_canceled)
```

**Cancellations and Market Segment**
```{r}
#Number of reservations by market segment
a = group_by(hotel_data, market_segment) %>%
  summarize(total_reservations = n())

#Number of uncanceled reservations by market segment
b = filter(hotel_data, is_canceled == 0)  %>%
group_by(market_segment) %>%
  summarize(number_uncanceled_reservations = n())

#Number of canceled reservations by market segment
c = filter(hotel_data, is_canceled == 1)  %>%
group_by(market_segment) %>%
  summarize(number_canceled_reservations = n())

reservations = merge(a, b, by="market_segment") %>%
  merge(c, by = "market_segment") %>%
  mutate(percent_canceled = number_canceled_reservations/total_reservations) %>%
  select(market_segment, percent_canceled, total_reservations, number_uncanceled_reservations, number_canceled_reservations)

reservations
```

**Average Daily Rate and Number of Adults**

We can see that as the number of adults in a booking increases through reservations of four adults, the average daily rate also increases. There are reservations with more than 4 people, but there are not enough to generate boxplots. 
```{r}
filter(hotel_data, adr < 1000, adults < 5) %>%
ggplot(aes(x = as.factor(adults), y = adr)) +
  geom_boxplot()
```

**Average Daily Rate and Month**
```{r}
#Average Daily Rate by Month
group_by(hotel_data, month_num) %>%
  summarize(avg_adr = mean(adr, na.rm = TRUE))
```

```{r}
ggplot(data = hotel_data, aes(x = month_num)) +
  geom_bar() +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1))
```




