---
title: "TO 414 Group Project 1"
author: "Shibi Sanjeev, Alexa Cornetta, Ben Tahmoush, Albi Popaj, Gjergj Camaj"
date: "2/20/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our first Group Project asks us to analyze the hotel_bookings data and extract meaningful business insights from it. We will walk you through our process and findings in this report. 

The data we will be working with can be found at: https://www.kaggle.com/jessemostipak/hotel-booking-demand


What role do meals play?
  
average daily rate by distribution channel



##Import
First, we are importing the hotel_bookings dataset, so that we may begin to explore and gain helpful insights from it. 

```{r}
hotel_data = read.csv("hotel_bookings.csv")

hotel_data$arrival_date_year = as.factor(hotel_data$arrival_date_year)
hotel_data$arrival_date_month = as.factor(hotel_data$arrival_date_month)
hotel_data$arrival_date_week_number = as.factor(hotel_data$arrival_date_week_number)
hotel_data$arrival_date_day_of_month = as.factor(hotel_data$arrival_date_day_of_month)
```

We will now install a package of helpful tools, called tidyverse, that we can use to manipulate and analyze the data we have imported. 

```{r}
library(tidyverse)
hotel_data$total_reservations = hotel_data$previous_bookings_not_canceled + hotel_data$previous_cancellations
```


## Make Factors 
Now we will identify variables that would be most helpful to us as factors, and convert them into such
```{r}
#summary(hotel_data)
#head(hotel_data)

# Factor needed variables
#hotel_data$is_canceled = as.factor(hotel_data$is_canceled)
hotel_data$is_repeated_guest = as.factor(hotel_data$is_repeated_guest)
hotel_data$company = as.factor(hotel_data$company)
hotel_data$hotel = as.factor(hotel_data$hotel)
hotel_data$arrival_date_month = as.factor(hotel_data$arrival_date_month)
hotel_data$country = as.factor(hotel_data$country)
hotel_data$market_segment = as.factor(hotel_data$market_segment)
hotel_data$distribution_channel = as.factor(hotel_data$distribution_channel)
hotel_data$reserved_room_type = as.factor(hotel_data$reserved_room_type)
hotel_data$assigned_room_type = as.factor(hotel_data$assigned_room_type)
hotel_data$agent = as.factor(hotel_data$agent)
hotel_data$reservation_status_date = as.factor(hotel_data$reservation_status_date)
hotel_data$reservation_status = as.factor(hotel_data$reservation_status)
hotel_data$customer_type = as.factor(hotel_data$customer_type)
hotel_data$deposit_type = as.factor(hotel_data$deposit_type)

```


##**Basic Data Structure**
To start, it will be helpful to get a good look at all of our data and how it is structured. We will do this below,
```{r}
#str(hotel_data)
```

### **Cancellations**

An important initial note. We don't have the data for every month in every year. In 2016, we have data for all twelve months. In 2017, we have data from January until August. In 2015, we only have data from July to December. The only months with three years of data are July and August. All other months only have two years worth of data from either 2015 and 2016 or 2016 and 2017.

Now we will begin exploring our data:

**Number of cancellations broken down by year and month**
```{r}
#This new column month_num has the month of the arrival at the hotel, similar to the arrival_date_month column. However, this one is in numerical form. This helps us with graphing. 
hotel_data$month_num = match(hotel_data$arrival_date_month, month.name)

#Number of cancellations broken down by year then month
res_breakdown_by_year_month = group_by(hotel_data, arrival_date_year, month_num, is_canceled) %>%
  summarize(num_reservations = n())

#Graphical representation of # of cancellations and reservations by year and month
ggplot(res_breakdown_by_year_month, aes(x = month_num, y = num_reservations)) +
  geom_bar(aes(fill = as.factor(is_canceled)),stat = "identity") +
  facet_wrap(~arrival_date_year) +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1)) +
  scale_fill_discrete(name = "", labels = c("Not Canceled", "Canceled")) +
  ggtitle("Number of Cancellations and Reservations in Each Year For Each Month") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Percent of Reservations Canceled in Each Year and Month**

When we break down the reservations by percentages, the actual percentage of reservations that gets canceled each month doesn't vary too much. It oscillates around about 30%. 

This number is more helpful for us to analyze, as conclusions could be skewed if we just look at the number of cancellations without taking into account the total reservations as well.
```{r}
#Number of reservations canceled by year, and month
t = hotel_data %>%
  group_by(arrival_date_year, month_num) %>%
  summarize(num_reservations_made = n()) %>%
   arrange(desc(arrival_date_year, month_num))

#Num reservations by year, and month
t1 = hotel_data %>%
  group_by(arrival_date_year, month_num, is_canceled) %>%
  summarize(num_reservations = n()) %>%
  arrange(arrival_date_year, month_num)

#Duplicating the dataframe t so that it can bind with dataframe t1
t2 = rbind(t, t) %>%
  arrange(arrival_date_year, month_num)

#Final Dataframe
t4 = (cbind(t1, t2) %>%
  select(arrival_date_year, month_num, is_canceled, num_reservations, num_reservations_made) %>%
  mutate(percent_reservations = num_reservations/num_reservations_made))

#Generating Bargraph
ggplot(t4, aes(x = month_num, y = percent_reservations)) +
  geom_bar(aes(fill = as.factor(is_canceled)),stat = "identity") +
  facet_wrap(~arrival_date_year) +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1)) +
  scale_y_continuous(name="Percent of Reservations") + 
  scale_fill_discrete(name = "", labels = c("Not Canceled", "Canceled")) +
  ggtitle("Percent of Reservations Canceled or Not Cancelled\n in Each Year For Each Month") +
  theme(plot.title = element_text(hjust = 0.5))
```


**Percent of Reservations Canceled in Each Year and Month Broken Down By Hotel Type**

When breaking down the cancellations further into the hotel type, we can see that, on average, trips to resorts are canceled less than trips to the city hotels. This is most likely because resort trips will more often than not be for vacation, whereas city hotel trips could be for business or a multitude of other reasons. An important note is that there almost twice as many reservations at city hotels than resorts, which could contribute to the difference in cancellation rates.


```{r}
#Number of reservations canceled by hotel type, year, and month
t1 = hotel_data %>%
  group_by(hotel, arrival_date_year, month_num, is_canceled) %>%
  summarize(num_reservations = n()) %>%
  arrange(hotel, arrival_date_year, month_num)

#Num reservations by hotel type, year, and month
t = hotel_data %>%
  group_by(hotel, arrival_date_year, month_num) %>%
  summarize(num_reservations_made = n()) %>%
   arrange(desc(hotel, arrival_date_year, month_num))

#Duplicating the dataframe t so that it can bind with dataframe t1
t2 = rbind(t, t) %>%
  arrange(hotel, arrival_date_year, month_num)

#Final Dataframe
t4 = (cbind(t1, t2) %>%
  select(hotel, arrival_date_year, month_num, is_canceled, num_reservations, num_reservations_made) %>%
  mutate(percent_reservations = num_reservations/num_reservations_made))

#Generating Bargraph
ggplot(t4, aes(x = month_num, y = percent_reservations)) +
  geom_bar(aes(fill = as.factor(is_canceled)),stat = "identity") +
  facet_grid(hotel ~ arrival_date_year) +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1)) +
  scale_y_continuous(name="Percent of Reservations") + 
  scale_fill_discrete(name = "", labels = c("Not Canceled", "Canceled")) +
  ggtitle("Percentage of Reservations Canceled or Not Canceled \n in Each Year For Each Month by Hotel") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Lead Time and Cancellations**
We can see that people who book earlier are more likely to cancel than those who book really late. This is likely because it is easier to plan for a shorter period of time and you are less likely to have unforeseen changes to their plans.
```{r}
#How many cancelations per days of lead time
group_by(hotel_data, is_canceled, lead_time) %>%
  summarize(count = n())

#Histograms of lead time by whether or not there were cancellations.
filter(hotel_data, lead_time < 600) %>%
ggplot(aes(x = lead_time)) +
  geom_histogram() +
  facet_wrap(~is_canceled)
```

## City vs Resort Hotels

We are looking now at if there is a meaningful difference between cancellations at city vs resort hotels
**Number of Reservations at City Hotels and Resort Hotels**
```{r}

group_by(hotel_data, hotel) %>%
  summarize(count = n(), percent_canceled = sum(is_canceled)/n(), avg_daily_rate = mean(adr, na.rm = T), avg_stay = mean(total_stay, na.rm = T))
```



**Length of Stay**
While the average stay is about the same length, resort hotels will have more longer stays than city hotels (they have a bigger range)

```{r}
#Calculating the total length of stay
hotel_data$total_stay = hotel_data$stays_in_week_nights + hotel_data$stays_in_weekend_nights

#Table of summary data of total stay broken down by city and resort hotels 
City_Hotel_Data = summary(filter(hotel_data, hotel == "City Hotel")$total_stay)

Resort_Hotel_Data = summary(filter(hotel_data, hotel == "Resort Hotel")$total_stay)

(total_stay_data = as.data.frame(cbind(City_Hotel_Data, Resort_Hotel_Data)))

filter(hotel_data, total_stay <= 20) %>%
ggplot(aes(x = hotel, y = total_stay)) +
  geom_boxplot() +
  ggtitle("Length of Stay Based on Hotel Type") +
  xlab("Hotel Type") +
  ylab("Length of Stay") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#Histogram showing distribution of percentage of reservations for length of stay broken down by city and resort hotels
filter(hotel_data, total_stay < 20) %>%
ggplot(aes(x = total_stay)) +
  geom_histogram(aes(y = stat(density)), binwidth = 1) +
  facet_wrap(~hotel) +
  ggtitle("Histogram of Length of Reservations for City and Resort Hotels") +
  scale_x_continuous(name="Number of Days") +
  scale_y_continuous(name="Percent of Reservations") +
  theme(plot.title = element_text(hjust = 0.5))
```


## **Market Segment**

**Market Segment and Average Daily Rate **

From this here, we can see that the average daily rate is highest when reservations are made with online travel agents, direct reservations, and aviation companies. Additionally, most reservations are made with online travel agents, offline travel agents or tour operators, direct reservations, and groups, respectively. 

An important note is that almost 61% of reservations made by groups are canceled. Another one is that while Aviation does have a high daily rate, there have only been 237 reservations placed over this two year period. 

We would recommend focusing marketing and trying to boost traction with online travel agents, direct reservations, offline travel agents or tour operators, and groups. This is because these all have relatively high average daily rates and make up a significant portion of the number of reservations. If you were to hone in on one, we would recommend focusing on online travel agents.
```{r}
(market_segment_info = (group_by(hotel_data, market_segment) %>%
  summarize(avg_daily_rate = round(mean(adr, na.rm = T), 2), avg_stay = mean(total_stay, na.rm = T), count = n(), num_cancellations = sum(is_canceled)) %>%
  mutate(percent_canceled = round((num_cancellations/count)*100, 2)) %>%
  arrange(desc(avg_daily_rate))))
```

```{r}
group_by(hotel_data, hotel, market_segment) %>%
  summarize(avg_daily_rate = round(mean(adr, na.rm = T)), count = n()) %>%
  arrange(market_segment)
```



**Cancellations and Market Segment**
Groups have the highest cancellation rate, but Online TAs have the highest number of cancelled reservations.
```{r}
#Number of reservations by market segment
a = group_by(hotel_data, market_segment) %>%
  summarize(total_reservations = n())

#Number of un-canceled reservations by market segment
b = filter(hotel_data, is_canceled == 0)  %>%
group_by(market_segment) %>%
  summarize(number_uncanceled_reservations = n())

#Number of canceled reservations by market segment
c = filter(hotel_data, is_canceled == 1)  %>%
group_by(market_segment) %>%
  summarize(number_canceled_reservations = n())

reservations = merge(a, b, by="market_segment") %>%
  merge(c, by = "market_segment") %>%
  mutate(percent_canceled = number_canceled_reservations/total_reservations) %>%
  select(market_segment, percent_canceled, total_reservations, number_uncanceled_reservations, number_canceled_reservations) %>%
  rename(
    number_uncanceled = number_uncanceled_reservations,
    number_canceled = number_canceled_reservations
  )

reservations
```

**Average Daily Rate and Number of Adults**
We can see that as the number of adults in a booking increases through reservations of four adults, the average daily rate also increases. There are reservations with more than 4 people, but there are not enough to generate boxplots. 
```{r}
filter(hotel_data, adr < 1000, adults < 5) %>%
ggplot(aes(x = as.factor(adults), y = adr)) +
  geom_boxplot()

filter(hotel_data, adr < 1000, children < 10) %>%
  ggplot(aes(x = as.factor(children), y = adr)) +
  geom_boxplot()
```

**Average Daily Rate and Month**
We can see from this that the month of August has the highest average daily rate.
```{r}
#Average Daily Rate by Month
group_by(hotel_data, month_num) %>%
  summarize(avg_adr = mean(adr, na.rm = TRUE))
```
##What is this?
```{r}
ggplot(data = hotel_data, aes(x = month_num)) +
  geom_bar() +
  scale_x_continuous(name="Month", breaks = seq(0,12, 1))
```

## Hotel By Country
```{r}
hotel_data$country = match(hotel_data$reservation_status,hotel_data$country )
num_cancellations_by_country = group_by(hotel_data, hotel_data$country, is_canceled)
summarize(num_cancellations_by_country)

```


```{r}
#ggplot(num_cancellations_by_country, aes(x = country, y = num_cancellations)) +
#  geom_bar(aes(fill = as.factor(is_canceled)),stat = "identity") +
#  facet_wrap(~arrival_date_year) +
#  scale_x_continuous(name="Country", breaks = seq(0,12, 1)) +
#  scale_fill_discrete(name = "", labels = c("Not Canceled", "Canceled")) +
#  ggtitle("Number of Cancellations and Reservations in Each Country") +
#  theme(plot.title = element_text(hjust = 0.5))
```
##Cancellations and Number of Children
We see here that having a higher number of children increases the chances of a cancellation. 
```{r}
causality(children, cause = NULL, vcov.=NULL, boot=FALSE, boot.runs=100)
```
##Cancellations and Repeated Guests
Here we will analyze the data to see if repeated guests are more likely to not cancel their reservation compared to non-repeat guests

This may lead us to the development of a loyalty program.
```{r}

```
##Deposit and Cancellations
Does requiring a deposit make a reservation less likely to be cancelled?
```{r}

```


##Conclusions
We should have a section here where we summarize all of the business insights we took away from this 






